---
title: "R Notebook"
---


# compute correlation matrices

The function to calculate correlations does the following:

- Filter species: Retains only species with at least 9 years of percent change data for either abundance or biomass.

- Initialize matrices: Creates two matrices, one for storing correlation values (corr_matrix) and one for storing the correlation type (type_matrix), for all pairs of species in the study.

- Loop through species pairs:
  For each pair of species, identifies common years for comparison.
  Extracts percent change vectors for both abundance and biomass for the overlapping years.

- Priority check:
  If both species have abundance data, computes the correlation for abundance changes.
  If both species only have biomass data, computes the correlation for biomass changes.
  If one species has abundance and the other has biomass, computes a mixed correlation.

- Store results: Saves the correlation value and the type (abundance, biomass, or mixed) in the matrices.
- Symmetry: Ensures the correlation matrix is symmetric by filling both upper and lower triangle values.


```{r}
library(dplyr)
library(sf)
library(ggplot2)
library(reshape2)
library(gridExtra)
library(tidyverse)
library(RColorBrewer)
library(ggplot2)
library(reshape2)
library(ggpubr)
#dataset_percent_change <- readRDS(here::here("processed_data/dataset_percent_change.rds"))
dataset_percent_change <- readRDS(here::here("output/temporary_output/dataset_percent_change_biome_genus.rds"))

```



# GENUS LEVEL + BIOME


```{r}

compute_correlations_biome <- function(data) {
  # Filter genera with at least 9 years of percent change in abundance or biomass
  filtered_data <- data %>%
    filter(!is.na(pct_change_abundance) | !is.na(pct_change_biomass)) %>%
    group_by(GENUS) %>%
    filter(n_distinct(YEAR) >= 6) %>%
    ungroup()

  # Check if there are at least two genera with sufficient data
  genera <- unique(filtered_data$GENUS)
  n_genera <- length(genera)
  if (n_genera < 2) {
    message("Not enough genera with sufficient data for correlation computation.")
    return(list(correlation = matrix(NA, n_genera, n_genera), type = matrix(NA, n_genera, n_genera)))
  }

  # Initialize correlation and type matrices
  corr_matrix <- matrix(NA, n_genera, n_genera)
  type_matrix <- matrix(NA, n_genera, n_genera)
  rownames(corr_matrix) <- genera
  colnames(corr_matrix) <- genera
  rownames(type_matrix) <- genera
  colnames(type_matrix) <- genera

  # Loop over genera pairs and compute correlations
  for (i in 1:(n_genera - 1)) {
    for (j in (i + 1):n_genera) {
      genus_i <- filtered_data %>% filter(GENUS == genera[i])
      genus_j <- filtered_data %>% filter(GENUS == genera[j])

      # Find overlapping years
      common_years <- intersect(genus_i$YEAR, genus_j$YEAR)

      # Ensure there are enough overlapping years for correlation
      if (length(common_years) < 2) {
        next  # Skip to the next pair if there are not enough overlapping years
      }

      # Extract percent change vectors for common years
      vec_i_abundance <- genus_i %>%
        filter(YEAR %in% common_years) %>%
        pull(pct_change_abundance)
      vec_i_biomass <- genus_i %>%
        filter(YEAR %in% common_years) %>%
        pull(pct_change_biomass)
      
      vec_j_abundance <- genus_j %>%
        filter(YEAR %in% common_years) %>%
        pull(pct_change_abundance)
      vec_j_biomass <- genus_j %>%
        filter(YEAR %in% common_years) %>%
        pull(pct_change_biomass)

      # Check if abundance is available for both genera and if vectors match in length
      if (!all(is.na(vec_i_abundance)) & !all(is.na(vec_j_abundance)) & length(vec_i_abundance) == length(vec_j_abundance)) {
        corr_matrix[i, j] <- cor(vec_i_abundance, vec_j_abundance, use = "complete.obs")
        type_matrix[i, j] <- "abundance"
      } else if (!all(is.na(vec_i_biomass)) & !all(is.na(vec_j_biomass)) & length(vec_i_biomass) == length(vec_j_biomass)) {
        corr_matrix[i, j] <- cor(vec_i_biomass, vec_j_biomass, use = "complete.obs")
        type_matrix[i, j] <- "biomass"
      } else {
        # Mixed correlation: one has abundance, the other has biomass
        if (!all(is.na(vec_i_abundance)) & !all(is.na(vec_j_biomass)) & length(vec_i_abundance) == length(vec_j_biomass)) {
          corr_matrix[i, j] <- cor(vec_i_abundance, vec_j_biomass, use = "complete.obs")
          type_matrix[i, j] <- "mixed"
        } else if (!all(is.na(vec_i_biomass)) & !all(is.na(vec_j_abundance)) & length(vec_i_biomass) == length(vec_j_abundance)) {
          corr_matrix[i, j] <- cor(vec_i_biomass, vec_j_abundance, use = "complete.obs")
          type_matrix[i, j] <- "mixed"
        }
      }

      # Fill the symmetric values
      corr_matrix[j, i] <- corr_matrix[i, j]
      type_matrix[j, i] <- type_matrix[i, j]
    }
  }

  return(list(correlation = corr_matrix, type = type_matrix))
}


```

```{r}
# Define the function to compute coherence matrices based on grouping criteria
compute_correlations_by_realm <- function(data, realm) {
  
  if (realm == "Terrestrial") {
    # Group by country and biome for terrestrial
    group_cols <- c("country", "BIOME_NAME")
  } else if (realm == "Marine") {
    # Group by biome only for marine
    group_cols <- c("marine_ecoregion")
  } else {
    stop("Invalid realm specified.")
  }
  
  # Define a list to store results
  realm_correlations <- list()
  
  # Loop through each unique combination in the grouping columns
  unique_combinations <- data %>%
    distinct(across(all_of(group_cols))) %>%
    drop_na()
  
  for (i in 1:nrow(unique_combinations)) {
    # Filter data for the specific combination
    specific_data <- data
    for (col in group_cols) {
      specific_data <- specific_data %>% filter(.data[[col]] == unique_combinations[[col]][i])
    }
    
    # Check if there's data for this combination
    if (nrow(specific_data) > 0) {
      # Compute correlations
      correlation_result <- compute_correlations_biome(specific_data)
      key <- paste(unique_combinations[i, , drop = TRUE], collapse = "_")
      realm_correlations[[key]] <- correlation_result
    }
  }
  
  return(realm_correlations)
}

# Split the data by realm
terrestrial_data <- dataset_percent_change %>% filter(REALM == "Terrestrial")
marine_data <- dataset_percent_change %>% filter(REALM == "Marine")

# Compute coherence matrices for terrestrial and marine realms
terrestrial_correlations <- compute_correlations_by_realm(terrestrial_data, "Terrestrial")
marine_correlations <- compute_correlations_by_realm(marine_data, "Marine")

```




# Plot EC patterns

```{r}
# Function to create a correlation matrix plot
plot_correlation_matrix <- function(correlation_matrix, title) {
  # Convert the matrix to long format for ggplot
  data_matrix <- melt(correlation_matrix, na.rm = TRUE)
  colnames(data_matrix) <- c("x", "y", "value")
  
  # Plot matrix with color scale
  palette <- RColorBrewer::brewer.pal(n = 11, name = "RdBu")
  ggplot(data_matrix, aes(x = x, y = y, fill = value)) +
    geom_raster() +
    scale_fill_gradientn(colors = rev(palette), limits = c(-1, 1), name = "Correlation") +
    labs(x = NULL, y = NULL, title = title) +
    theme_classic() +
    theme(
      legend.position = "right",
      axis.text = element_blank(),
      axis.title = element_blank(),
      axis.ticks = element_blank(),
      axis.line = element_blank(),
      plot.title = element_text(hjust = 0.5)
    ) +
    coord_fixed()
}

# Function to create a distribution plot of correlation values
plot_distribution <- function(correlation_matrix, title) {
  # Extract correlation values and remove NAs
  cor_values <- correlation_matrix[lower.tri(correlation_matrix, diag = FALSE)]
  cor_values <- cor_values[!is.na(cor_values)]
  
  # Bin the correlation values
  data_new <- data.frame(correlation = cor_values) %>%
    dplyr::mutate(cor_bin = cut(correlation, breaks = seq(-0.9, 0.9, by = 0.1), include.lowest = TRUE)) %>%
    dplyr::group_by(cor_bin) %>%
    dplyr::summarize(frequency = n()) %>%
    dplyr::mutate(cor_bin_mid = (as.numeric(sub("\\[(.+),.+\\)", "\\1", cor_bin)) + 
                                 as.numeric(sub(".+,(.+)\\)", "\\1", cor_bin))) / 2)
  
  # Plot distribution with standardized color scale
  ggplot(data_new, aes(x = cor_bin, y = frequency, fill = cor_bin_mid)) +
    geom_bar(stat = "identity") +
    scale_fill_distiller(palette = "RdBu", limits = c(-0.8, 0.8), direction = -1, name = "Correlation") +
    labs(x = "Interacting species co-response", y = "Frequency", title = title) +
    theme_classic() +
    theme(
      legend.position = "none",
      plot.title = element_text(hjust = 0.5),
      axis.text.x = element_text(angle = 45, hjust = 1)
    )
}
```




```{r}
# Initialize lists to store the plots for terrestrial and marine assemblages
terrestrial_matrix_plots <- list()
terrestrial_distribution_plots <- list()
marine_matrix_plots <- list()
marine_distribution_plots <- list()

# Generate terrestrial plots
for (assemblage_name in names(terrestrial_correlations)) {
  correlation_data <- terrestrial_correlations[[assemblage_name]]
  if (!is.null(correlation_data$correlation) && nrow(correlation_data$correlation) > 0) {
    # Create matrix plot and distribution plot
    terrestrial_matrix_plots[[assemblage_name]] <- plot_correlation_matrix(
      correlation_data$correlation, title = paste("Terrestrial:", assemblage_name)
    )
    terrestrial_distribution_plots[[assemblage_name]] <- plot_distribution(
      correlation_data$correlation, title = paste("Terrestrial:", assemblage_name)
    )
  }
}

# Generate marine plots
for (assemblage_name in names(marine_correlations)) {
  correlation_data <- marine_correlations[[assemblage_name]]
  if (!is.null(correlation_data$correlation) && nrow(correlation_data$correlation) > 0) {
    # Create matrix plot and distribution plot
    marine_matrix_plots[[assemblage_name]] <- plot_correlation_matrix(
      correlation_data$correlation, title = paste("Marine:", assemblage_name)
    )
    marine_distribution_plots[[assemblage_name]] <- plot_distribution(
      correlation_data$correlation, title = paste("Marine:", assemblage_name)
    )
  }
}




```



```{r}
# Arrange terrestrial matrix plots
ggarrange(plotlist = terrestrial_matrix_plots, ncol = 3, nrow = 4, common.legend = TRUE)
# Arrange terrestrial distribution plots
ggarrange(plotlist = terrestrial_distribution_plots, ncol = 3, nrow = 4, common.legend = TRUE)

# Arrange marine matrix plots
ggarrange(plotlist = marine_matrix_plots, ncol = 3, nrow = 3, common.legend = TRUE)
# Arrange marine distribution plots
ggarrange(plotlist = marine_distribution_plots, ncol = 3, nrow = 3, common.legend = TRUE)
```

