---
title: "Compute marine ecoregions"
---


```{r}
library(tidyverse)
library(dplyr)
library(ggpubr)
library(ggplot2)
library(geosphere)
library(sf)
library(igraph)
library(rnaturalearth)
# Load world map
world <- ne_countries(scale = "medium", returnclass = "sf")
```

```{r}
# # INPUT FILES #
biotime.raw <- read.csv(here::here('../BioTIMEQuery_24_06_2021.csv'))
metadata <- read.csv(here::here('raw_data/BioTIMEMetadata_24_06_2021.csv'))
data_with_biome <- readRDS(here::here("output/temporary_output/data_with_biomes.rds"))
marine_data_with_ecoregions <- readRDS(here::here("output/marine_data_with_ecoregions.rds"))

```



```{r}
# Step 2: Extract unique terrestrial coordinates
marine_studies <- metadata %>%
  filter(REALM == "Marine") %>%
  pull(STUDY_ID)

unique_coords_marine <- biotime.raw %>%
  filter(STUDY_ID %in% marine_studies, !is.na(LATITUDE), !is.na(LONGITUDE)) %>%
  distinct(LATITUDE, LONGITUDE) %>%
  st_as_sf(coords = c("LONGITUDE", "LATITUDE"), crs = 4326)
```



```{r}
# marine_data <- data_with_biome %>%
#   filter(REALM == "Marine")
```



```{r}

# Replace 'path_to_shapefile' with the actual path to the downloaded MEOW shapefile
meow_shapefile <- st_read(here::here("../marine_regions/meow_ecos.shp"))

st_crs(meow_shapefile)

# Load the world map as background
world <- rnaturalearth::ne_countries(scale = "medium", returnclass = "sf")

# Plot the marine ecoregions with distinct colors by `REALM`
ggplot() +
  geom_sf(data = world, fill = "gray80", color = "black", size = 0.1) +
  geom_sf(data = meow_shapefile, aes(fill = REALM), color = NA) +
  labs(title = "Marine Ecoregions by Realm", fill = "Realm") +
  theme_minimal()
```



```{r}
# Check and align CRS
# Confirm the CRS for `marine_data` and `meow_shapefile`
crs_marine_data <- st_crs(unique_coords_marine)
crs_meow_shapefile <- st_crs(meow_shapefile)

if (crs_marine_data != crs_meow_shapefile) {
  marine_data <- st_transform(unique_coords_marine, crs = st_crs(meow_shapefile))
}

# Perform the spatial join
marine_data_with_ecoregions <- st_join(unique_coords_marine, meow_shapefile[, c("ECOREGION", "PROVINCE", "REALM")], left = TRUE)

marine_data_with_ecoregions <- marine_data_with_ecoregions |> 
  rename(marine_ecoregion = REALM)

head(marine_data_with_ecoregions)

```


plot coordinates and ecoregions assigned

```{r}
ggplot(data = world) +
  geom_sf(fill = "gray90", color = "black") +
  geom_sf(data = marine_data_with_ecoregions, aes(color = marine_ecoregion), size = 1, alpha = 0.7) +
  labs(title = "Marine Data Points Colored by marine ecoregion",
       subtitle = "Each point represents a data point colored by its assigned marine ecoregion",
       color = "Marine ecoregion",
       x = "Longitude", y = "Latitude") +
  theme_minimal() +
  theme(legend.position = "right")
```



# Manual checks to add marine ecoregion to those coordinates without it


plotting points without marine ecoregions assigned

```{r}
# Filter points with missing ecoregion information
missing_ecoregion_points <- marine_data_with_ecoregions %>%
  filter(is.na(marine_ecoregion))

# Plot the world map with points that have missing ecoregion information
ggplot(data = world) +
  geom_sf(fill = "gray90", color = "black") +
  geom_sf(data = missing_ecoregion_points, color = "red", size = 1, alpha = 0.7) +
  labs(title = "Marine Data Points Missing Ecoregion Information",
       subtitle = "Red points indicate missing marine ecoregion assignment",
       x = "Longitude", y = "Latitude") +
  theme_minimal()
```


Plot the first missing point/s


```{r}
# Select a specific point by index (e.g., first point in missing_location_points)
single_point <- missing_ecoregion_points[1:15, ]

# Plot the selected point on the map
ggplot(data = world) +
  geom_sf(fill = "gray90", color = "black") +
  geom_sf(data = single_point, color = "red", size = 3) +
  labs(title = "Inspecting a Single Point for Country Assignment") +
  theme_minimal()

```


```{r}
unique(marine_data_with_ecoregions$marine_ecoregion)
```



Update the data with verified ecoregion

```{r}
# Define the range of indices you want to update and the new ecoregion name
start_index <- 1  # First point to update
end_index <- 15    # Last point to update
ecoregion_name <- "Temperate Australasia"  # Verified ecoregion name for these points

# Extract the coordinates of the points in the selected range
latitudes <- st_coordinates(missing_ecoregion_points[start_index:end_index, ])[,2]
longitudes <- st_coordinates(missing_ecoregion_points[start_index:end_index, ])[,1]

# Update the `marine_ecoregion` field in `marine_data_with_ecoregions` based on matching coordinates
marine_data_with_ecoregions <- marine_data_with_ecoregions %>%
  mutate(
    marine_ecoregion = if_else(
      st_coordinates(.)[,2] %in% latitudes & st_coordinates(.)[,1] %in% longitudes,
      ecoregion_name,
      marine_ecoregion
    )
  )

#saveRDS(marine_data_with_ecoregions, here::here("output/temporary_output/marine_data_with_ecoregions.rds"))

#marine_data_with_ecoregions <- readRDS(here::here("output/temporary_output/marine_data_with_ecoregions.rds"))


```


```{r}
# Assign 'Southern Ocean' to points south of 60°S
marine_data_with_ecoregions <- marine_data_with_ecoregions %>%
  mutate(
    marine_ecoregion = if_else(
      is.na(marine_ecoregion) & st_coordinates(.)[,2] < -60,
      "Southern Ocean",
      marine_ecoregion
    )
  )

# Filter points south of 60°S
southern_ocean_points <- marine_data_with_ecoregions1 %>%
  filter(st_coordinates(geometry)[, 2] < -60)

# Plot the world map with Southern Ocean points
ggplot(data = world) +
  geom_sf(fill = "gray90", color = "black") +
  geom_sf(data = southern_ocean_points, color = "blue", size = 1, alpha = 0.7) +
  labs(title = "Marine Data Points in the Southern Ocean",
       subtitle = "Points south of 60°S assigned to the Southern Ocean",
       x = "Longitude", y = "Latitude") +
  theme_minimal()

```



------------------------------------------

```{r}
# Validate geometries
missing_ecoregion_points <- st_make_valid(missing_ecoregion_points)
grid_sf <- st_make_valid(grid_sf)
```


# Update marine ecoregion based on grid

```{r}
ggplot(data = world) +
  geom_sf(fill = "gray90", color = "black") +  # Base world map
  geom_sf(data = marine_data_with_ecoregions, aes(color = marine_ecoregion), size = 1, alpha = 0.7) +  # Marine data points
  scale_color_viridis_d(name = "Marine Ecoregion", option = "C", na.value = "grey50") +  # Viridis color palette for marine ecoregions
  labs(
    title = "Marine Data Points by Marine Ecoregion",
    x = "Longitude", y = "Latitude"
  ) +
  theme_minimal() +
  theme(
    legend.position = "right",
    legend.title = element_text(size = 10),
    legend.text = element_text(size = 8)
  )
```



plotting points without marine ecoregions assigned

```{r}
# Filter points with missing ecoregion information
missing_ecoregion_points <- marine_data_with_ecoregions %>%
  filter(is.na(marine_ecoregion))

# Plot the world map with points that have missing ecoregion information
ggplot(data = world) +
  geom_sf(fill = "gray90", color = "black") +
  geom_sf(data = missing_ecoregion_points, color = "red", size = 1, alpha = 0.7) +
  labs(title = "Marine Data Points Missing Ecoregion Information",
       subtitle = "Red points indicate missing marine ecoregion assignment",
       x = "Longitude", y = "Latitude") +
  theme_minimal()
```



```{r}
# Define grid parameters (adjust cell size if necessary)
cell_size <- 10  # Cell size in degrees
bounding_box <- st_bbox(missing_ecoregion_points)  # Bounding box for missing points
origin <- c(bounding_box$xmin, bounding_box$ymin)  # Lower-left corner of the grid

# Create the grid
grid <- st_make_grid(
  st_as_sfc(bounding_box),
  cellsize = cell_size,
  offset = origin,
  crs = st_crs(missing_ecoregion_points)
)
grid_sf <- st_sf(grid_id = 1:length(grid), geometry = grid)

# Define the grid IDs or ranges you want to highlight
# Example: select grid cells with IDs from 10 to 20 on the x-axis and 5 to 15 on the y-axis
highlighted_cells <- grid_sf %>%
  filter(
    grid_id >= 1 & grid_id <= 36 |  # Example x-range
    grid_id >= 1 & grid_id <= 4     # Example y-range
  )

# Plot the map with grid cells and highlight selected cells
ggplot(data = world) +
  geom_sf(fill = "gray90", color = "black") +
  geom_sf(data = grid_sf, fill = NA, color = "blue", linetype = "dashed") +  # Outline all grid cells
  geom_sf(data = highlighted_cells, fill = "yellow", color = "blue", alpha = 0.5) +  # Highlighted cells
  geom_sf(data = missing_ecoregion_points, color = "red", size = 1, alpha = 0.7) +
  labs(
    title = "Marine Data Points Missing Ecoregion Information",
    subtitle = "Highlighted Grid Cells",
    x = "Longitude", y = "Latitude"
  ) +
  theme_minimal()

```

```{r}
# Define the cell size for grid cells, in degrees
cell_size <- 10  # degrees, adjust as needed

# Create the grid based on geographic coordinates
grid <- st_make_grid(
  st_as_sfc(st_bbox(missing_ecoregion_points)),
  cellsize = c(cell_size, cell_size),
  crs = st_crs(missing_ecoregion_points)
)

# Convert the grid to an sf object for labeling or further processing
grid_sf <- st_sf(grid_id = 1:length(grid), geometry = grid)

# Plot the world map with parallels and meridians
ggplot(data = world) +
  geom_sf(fill = "gray90", color = "black") +
  geom_sf(data = grid_sf, fill = NA, color = "blue", linetype = "dashed") +  # Plot grid cells
  geom_sf(data = missing_ecoregion_points, color = "red", size = 1, alpha = 0.7) +
  coord_sf(expand = FALSE, datum = sf::st_crs(4326)) +  # Keep it in latitude/longitude coordinates
  scale_x_continuous(breaks = seq(-180, 180, by = cell_size)) +  # Adds meridians at intervals
  scale_y_continuous(breaks = seq(-90, 90, by = cell_size)) +    # Adds parallels at intervals
  labs(
    title = "Marine Data Points with Missing Ecoregion Information",
    subtitle = "Grid cells based on latitude/longitude",
    x = "Longitude", y = "Latitude"
  ) +
  theme_minimal()

```

```{r}

lon_min <- -180  # Adjust these values based on the plot
lon_max <- -120
lat_min <- -30
lat_max <- 0

# Step 1: Transform grid_sf to a planar projection suitable for centroids
grid_sf_projected <- st_transform(grid_sf, crs = 3395)  # World Mercator projection

# Step 2: Calculate centroids in the projected CRS
grid_sf_projected$centroid <- st_centroid(grid_sf_projected$geometry)

# Step 3: Transform centroids back to the original CRS
grid_sf$centroid <- st_transform(grid_sf_projected$centroid, crs = st_crs(grid_sf))

# Step 4: Extract coordinates from centroids and filter for selected cells
centroid_coords <- st_coordinates(grid_sf$centroid)
selected_grid_cells <- grid_sf %>%
  filter(
    centroid_coords[,1] >= lon_min &
    centroid_coords[,1] <= lon_max &
    centroid_coords[,2] >= lat_min &
    centroid_coords[,2] <= lat_max
  )

# Plot with the selected grid cells highlighted in yellow
ggplot(data = world) +
  geom_sf(fill = "gray90", color = "black") +
  geom_sf(data = grid_sf, fill = NA, color = "blue", linetype = "dashed") +  # Plot all grid cells
  geom_sf(data = selected_grid_cells, fill = "yellow", color = "blue", linetype = "dashed") +  # Highlight selected cells in yellow
  geom_sf(data = missing_ecoregion_points, color = "red", size = 1, alpha = 0.7) +
  coord_sf(expand = FALSE, datum = sf::st_crs(4326)) +  # Keep it in latitude/longitude coordinates
  scale_x_continuous(breaks = seq(-180, 180, by = cell_size)) +  # Adds meridians at intervals
  scale_y_continuous(breaks = seq(-90, 90, by = cell_size)) +    # Adds parallels at intervals
  labs(
    title = "Marine Data Points with Missing Ecoregion Information",
    subtitle = "Selected grid cells highlighted in yellow",
    x = "Longitude", y = "Latitude"
  ) +
  theme_minimal()


```
```{r}
unique(marine_data_with_ecoregions$marine_ecoregion)
```


```{r}
# Filter `missing_ecoregion_points` within these boundaries
selected_points <- missing_ecoregion_points %>%
  filter(
    st_coordinates(.)[,1] >= lon_min & st_coordinates(.)[,1] <= lon_max &
    st_coordinates(.)[,2] >= lat_min & st_coordinates(.)[,2] <= lat_max
  )

# Update `marine_ecoregion` for these points to "Southern Ocean" (for example)
marine_data_with_ecoregions <- marine_data_with_ecoregions %>%
  mutate(
    marine_ecoregion = if_else(
      st_coordinates(.)[,1] >= lon_min & st_coordinates(.)[,1] <= lon_max &
      st_coordinates(.)[,2] >= lat_min & st_coordinates(.)[,2] <= lat_max,
      "Eastern Indo-Pacific",
      marine_ecoregion
    )
  )

```

```{r}
head(marine_data_with_ecoregions)

```

---------------------------

Fine tunning ecoregions

- Temperate Northern Atlantic -> divided into American and European, Mediterranean Sea, and Black Sea

- Western Indo-Pacific and Australasia expanded south, until latitude 60 where the South Sea starts


```{r}
ggplot(data = world) +
  geom_sf(fill = "gray90", color = "black") +  # Base world map
  geom_sf(data = marine_data_with_ecoregions, aes(color = marine_ecoregion), size = 1, alpha = 0.7) +  # Marine data points
  #scale_color_viridis_d(name = "Marine Ecoregion", option = "turbo", na.value = "grey50") +  # Viridis color palette for marine ecoregions
  labs(
    title = "Marine Data Points by Marine Ecoregion",
    x = "Longitude", y = "Latitude"
  ) +
  theme_minimal() +
  theme(
    legend.position = "right",
    legend.title = element_text(size = 10),
    legend.text = element_text(size = 8)
  )

#ggsave(here::here("output/marine_ecoregions.png"), height = 10, width = 13)

```

```{r}
unique(marine_data_with_ecoregions$marine_ecoregion)
```
```{r}
# Define the cell size for grid cells, in degrees
cell_size <- 10  # degrees, adjust as needed

# Create the grid based on geographic coordinates
grid <- st_make_grid(
  st_as_sfc(st_bbox(marine_data_with_ecoregions)),
  cellsize = c(cell_size, cell_size),
  crs = st_crs(missing_ecoregion_points)
)

# Convert the grid to an sf object for labeling or further processing
grid_sf <- st_sf(grid_id = 1:length(grid), geometry = grid)
```



```{r}

lon_min <- -70  # Adjust these values based on the plot
lon_max <- -40
lat_min <- 55
lat_max <- 60

# Step 1: Transform grid_sf to a planar projection suitable for centroids
grid_sf_projected <- st_transform(grid_sf, crs = 3395)  # World Mercator projection

# Step 2: Calculate centroids in the projected CRS
grid_sf_projected$centroid <- st_centroid(grid_sf_projected$geometry)

# Step 3: Transform centroids back to the original CRS
grid_sf$centroid <- st_transform(grid_sf_projected$centroid, crs = st_crs(grid_sf))

# Step 4: Extract coordinates from centroids and filter for selected cells
centroid_coords <- st_coordinates(grid_sf$centroid)
selected_grid_cells <- grid_sf %>%
  filter(
    centroid_coords[,1] >= lon_min &
    centroid_coords[,1] <= lon_max &
    centroid_coords[,2] >= lat_min &
    centroid_coords[,2] <= lat_max
  )

# Plot with the selected grid cells highlighted in yellow
ggplot(data = world) +
  geom_sf(fill = "gray90", color = "black") +
  geom_sf(data = grid_sf, fill = NA, color = "blue", linetype = "dashed") +  # Plot all grid cells
  geom_sf(data = selected_grid_cells, fill = "yellow", color = "blue", linetype = "dashed") +  # Highlight selected cells in yellow
  geom_sf(data = marine_data_with_ecoregions, aes(color = marine_ecoregion), size = 1, alpha = 0.7) +
  coord_sf(expand = FALSE, datum = sf::st_crs(4326)) +  # Keep it in latitude/longitude coordinates
  scale_x_continuous(breaks = seq(-180, 180, by = cell_size)) +  # Adds meridians at intervals
  scale_y_continuous(breaks = seq(-90, 90, by = cell_size)) +    # Adds parallels at intervals
  labs(
    title = "Marine Ecoregions and Grid System for Fine Tunning Point Assignment",
    subtitle = "Selected grid cells highlighted in yellow",
    x = "Longitude", y = "Latitude"
  ) +
  theme_minimal()

#ggsave(here::here("output/temporary_output/grid_system_ecoregion_assingment.png"), height = 10, width = 13)

```

```{r}
# Filter `missing_ecoregion_points` within these boundaries
selected_points <- marine_data_with_ecoregions %>%
  filter(
    st_coordinates(.)[,1] >= lon_min & st_coordinates(.)[,1] <= lon_max &
    st_coordinates(.)[,2] >= lat_min & st_coordinates(.)[,2] <= lat_max
  )

# Update `marine_ecoregion` for these points to "Southern Ocean" (for example)
marine_data_with_ecoregions <- marine_data_with_ecoregions %>%
  mutate(
    marine_ecoregion = if_else(
      st_coordinates(.)[,1] >= lon_min & st_coordinates(.)[,1] <= lon_max &
      st_coordinates(.)[,2] >= lat_min & st_coordinates(.)[,2] <= lat_max,
      "Greenland Sea",
      marine_ecoregion
    )
  )

```


```{r}
# Update the ecoregion name
marine_data_with_ecoregions <- marine_data_with_ecoregions %>%
  mutate(marine_ecoregion = recode(marine_ecoregion,
                                   "Arctic" = "Arctic sea"))
```

```{r}
#saveRDS(marine_data_with_ecoregions, here::here("output/marine_data_with_ecoregions.rds"))
```

