---
title: "Compute proportion change and correlation matrix from raw data"
Authors: AF
---


*INPUT*: raw data
*PROCESS*: 
(1) convert 0 for abundance or biomass to NA to not consider it for calculating % change
(2) compute % change in abundance and biomass
(3) compute correlation matrix for a single study as an example - ready to make ir for all
*OUTPUT*:


```{r}
library(tidyverse)
library(dplyr)
library(ggpubr)
library(ggplot2)
library(reshape2)
```

# DATA PROCESSING 


```{r}
# # INPUT FILES # #
biotime.raw <- read.csv(here::here('../BioTIMEQuery_24_06_2021.csv'))
metadata <- read.csv(here::here('data/prep_biotime/BioTIMEMetadata_24_06_2021.csv'))

biotime_subset <- read.csv(here::here('AF_pilot/processed_data/subset_abundance_only.csv'))
```


- Remove all 0 or NA for biomass or abundance
- Aggregate the plot level data of species to study level - mean among plots per study

```{r}
# Step 0: Create a unique species name in the form "species_genus" at the beginning
biotime.raw$SPECIES_GENUS <- paste(biotime.raw$SPECIES, biotime.raw$GENUS, sep = "_")

# Step 1: Handle potential false 0s in abundance and biomass
# Replace 0s with NAs if the study didn't record how abundance or biomass was measured
biotime.raw$sum.allrawdata.ABUNDANCE[which(
  biotime.raw$STUDY_ID %in% 
    metadata$STUDY_ID[which(is.na(metadata$ABUNDANCE_TYPE))])] <- NA

biotime.raw$sum.allrawdata.BIOMASS[which(
  biotime.raw$STUDY_ID %in% 
    metadata$STUDY_ID[which(is.na(metadata$BIOMASS_TYPE))])] <- NA

# Step 2: Handle missing plot information by replacing NAs in PLOT with a placeholder
biotime.raw$PLOT[which(is.na(biotime.raw$PLOT))] <- "Was_NA"


# Step 3: Create a UNIQUE_ID to aggregate data per study, species_genus, and year across all plots
biotime.raw$UNIQUE_ID <- paste(biotime.raw$STUDY_ID, biotime.raw$SPECIES_GENUS, biotime.raw$YEAR, sep = "~")

# Step 4: Aggregate both abundance and biomass data by study, species_genus, and year across all plots
# We aggregate abundance and biomass together and handle NAs automatically
biotime.agg <- aggregate(cbind(sum.allrawdata.ABUNDANCE, sum.allrawdata.BIOMASS) ~ UNIQUE_ID, 
                         data = biotime.raw, 
                         FUN = function(x) if (all(is.na(x))) NA else mean(x, na.rm = TRUE))

# Step 5: Add identifying columns (STUDY_ID, SPECIES_GENUS, YEAR) back into the aggregated dataset
biotime.agg$STUDY_ID <- biotime.raw$STUDY_ID[match(biotime.agg$UNIQUE_ID, biotime.raw$UNIQUE_ID)]
biotime.agg$SPECIES_GENUS <- biotime.raw$SPECIES_GENUS[match(biotime.agg$UNIQUE_ID, biotime.raw$UNIQUE_ID)]
biotime.agg$YEAR <- biotime.raw$YEAR[match(biotime.agg$UNIQUE_ID, biotime.raw$UNIQUE_ID)]

# Step 6: Calculate the sample size (number of plots) that went into calculating the mean for both abundance and biomass
ss_abun <- as.data.frame(table(biotime.raw$UNIQUE_ID[-which(is.na(biotime.raw$sum.allrawdata.ABUNDANCE))]))
ss_bio <- as.data.frame(table(biotime.raw$UNIQUE_ID[-which(is.na(biotime.raw$sum.allrawdata.BIOMASS))]))

# Merge sample size information into the aggregated dataset
biotime.agg$SAMPLE_SIZE_ABUNDANCE <- ss_abun$Freq[match(biotime.agg$UNIQUE_ID, ss_abun$Var1)]
biotime.agg$SAMPLE_SIZE_BIOMASS <- ss_bio$Freq[match(biotime.agg$UNIQUE_ID, ss_bio$Var1)]

# Step 7: Remove rows where both abundance and biomass are zero, or one is zero and the other is NA
biotime.agg <- biotime.agg %>%
  dplyr::filter(!(sum.allrawdata.ABUNDANCE == 0 & (sum.allrawdata.BIOMASS == 0 | is.na(sum.allrawdata.BIOMASS)))) %>%
  dplyr::filter(!(sum.allrawdata.BIOMASS == 0 & (sum.allrawdata.ABUNDANCE == 0 | is.na(sum.allrawdata.ABUNDANCE))))

# Step 8: Create a new variable for study and species_genus
biotime.agg$STUDY_SPECIES_GENUS <- paste(biotime.agg$STUDY_ID, biotime.agg$SPECIES_GENUS, sep = "~")


head(biotime.agg)
```



# compute % change in abundance or biomass

```{r}
# Step 1: Sort the data by STUDY_SPECIES_GENUS and YEAR
biotime.agg <- biotime.agg %>%
  arrange(STUDY_SPECIES_GENUS, YEAR)

# Step 2: Compute the percentage change for abundance and biomass from year t to year t+1
biotime.agg <- biotime.agg %>%
  group_by(STUDY_SPECIES_GENUS) %>%  # Group by species_genus within each study
  mutate(
    pct_change_abundance = (sum.allrawdata.ABUNDANCE - lag(sum.allrawdata.ABUNDANCE)) / lag(sum.allrawdata.ABUNDANCE) * 100,
    pct_change_biomass = (sum.allrawdata.BIOMASS - lag(sum.allrawdata.BIOMASS)) / lag(sum.allrawdata.BIOMASS) * 100
  ) %>%
  ungroup()  # Ungroup the data after calculation


head(biotime.agg)
```

Check:

```{r}
# Check if any species has NA for the percentage change in abundance but a valid value for the biomass change:
# Check for species with NA for abundance change but a valid biomass change
check_abundance_na_biomass_valid <- biotime.agg %>%
  filter(is.na(pct_change_abundance) & !is.na(pct_change_biomass))

head(check_abundance_na_biomass_valid)
```



# compute correlation matrices

The function to calculate correlations does the following:

- Filter species: Retains only species with at least 9 years of percent change data for either abundance or biomass.

- Initialize matrices: Creates two matrices, one for storing correlation values (corr_matrix) and one for storing the correlation type (type_matrix), for all pairs of species in the study.

- Loop through species pairs:
  For each pair of species, identifies common years for comparison.
  Extracts percent change vectors for both abundance and biomass for the overlapping years.

- Priority check:
  If both species have abundance data, computes the correlation for abundance changes.
  If both species only have biomass data, computes the correlation for biomass changes.
  If one species has abundance and the other has biomass, computes a mixed correlation.

- Store results: Saves the correlation value and the type (abundance, biomass, or mixed) in the matrices.
- Symmetry: Ensures the correlation matrix is symmetric by filling both upper and lower triangle values.


```{r}
# Function to calculate correlations for a given STUDY_ID
compute_correlations <- function(data) {
  # Filter species with at least 9 years of percent change in abundance or biomass
  filtered_data <- data %>%
    filter(!is.na(pct_change_abundance) | !is.na(pct_change_biomass)) %>%
    group_by(SPECIES_GENUS) %>%
    filter(n() >= 9) %>%
    ungroup()

  # Initialize correlation matrix and type matrix
  species <- unique(filtered_data$SPECIES_GENUS)
  n_species <- length(species)
  corr_matrix <- matrix(NA, n_species, n_species)
  type_matrix <- matrix(NA, n_species, n_species)
  rownames(corr_matrix) <- species
  colnames(corr_matrix) <- species
  rownames(type_matrix) <- species
  colnames(type_matrix) <- species

  # Loop over species pairs and compute correlations
  for (i in 1:(n_species - 1)) {
    for (j in (i + 1):n_species) {
      species_i <- filtered_data %>%
        filter(SPECIES_GENUS == species[i])
      species_j <- filtered_data %>%
        filter(SPECIES_GENUS == species[j])

      # Find overlapping years
      common_years <- intersect(species_i$YEAR, species_j$YEAR)

      # Extract the percent change vectors for common years
      vec_i_abundance <- species_i %>%
        filter(YEAR %in% common_years) %>%
        pull(pct_change_abundance)
      vec_i_biomass <- species_i %>%
        filter(YEAR %in% common_years) %>%
        pull(pct_change_biomass)
      
      vec_j_abundance <- species_j %>%
        filter(YEAR %in% common_years) %>%
        pull(pct_change_abundance)
      vec_j_biomass <- species_j %>%
        filter(YEAR %in% common_years) %>%
        pull(pct_change_biomass)

      # Check if abundance is available for both species
      if (!all(is.na(vec_i_abundance)) & !all(is.na(vec_j_abundance))) {
        # Use abundance for both species
        corr_matrix[i, j] <- cor(vec_i_abundance, vec_j_abundance, use = "complete.obs")
        type_matrix[i, j] <- "abundance"
      } else if (!all(is.na(vec_i_biomass)) & !all(is.na(vec_j_biomass))) {
        # Use biomass for both species
        corr_matrix[i, j] <- cor(vec_i_biomass, vec_j_biomass, use = "complete.obs")
        type_matrix[i, j] <- "biomass"
      } else {
        # One species has abundance, the other has biomass (mixed correlation)
        if (!all(is.na(vec_i_abundance)) & !all(is.na(vec_j_biomass))) {
          corr_matrix[i, j] <- cor(vec_i_abundance, vec_j_biomass, use = "complete.obs")
          type_matrix[i, j] <- "mixed"
        } else if (!all(is.na(vec_i_biomass)) & !all(is.na(vec_j_abundance))) {
          corr_matrix[i, j] <- cor(vec_i_biomass, vec_j_abundance, use = "complete.obs")
          type_matrix[i, j] <- "mixed"
        }
      }

      # Fill the symmetric values
      corr_matrix[j, i] <- corr_matrix[i, j]
      type_matrix[j, i] <- type_matrix[i, j]
    }
  }

  return(list(correlation = corr_matrix, type = type_matrix))
}
```

- Testing for one study only

```{r}
# Select a specific STUDY_ID (for example, STUDY_ID = 119)
specific_study_data <- biotime.agg %>%
  filter(STUDY_ID == 119)

# Apply the compute_correlations function to this specific study
correlation_result <- compute_correlations(specific_study_data)

# View the correlation matrix and the type matrix for this specific study
correlation_matrix <- correlation_result$correlation
type_matrix <- correlation_result$type

# Display the results
print("Correlation Matrix:")
print(correlation_matrix)

print("Type Matrix (Abundance, Biomass, or Mixed):")
print(type_matrix)
```


Plot of that one matrix

```{r}
data_matrix <- melt(correlation_matrix, na.rm = TRUE)
colnames(data_matrix) <- c("x", "y", "value")

# Plot the matrix
palette <- RColorBrewer::brewer.pal(n = 11, name = "RdBu")

plot_matrix_c <- ggplot(data_matrix, aes(x = x, y = y, fill = value)) +
  geom_raster() +
  scale_fill_gradientn(colors = rev(palette), limits = c(-1, 1)) +
  labs(x = NULL, y = NULL, title = expression("Co-response matrix"~italic("C"))) +
  theme_classic() +
  theme(
    legend.position = "right",
    axis.text = element_blank(),
    axis.title = element_blank(),
    axis.ticks = element_blank(),  
    axis.line = element_blank(),   
    plot.title = element_text(hjust = 0.5)
  ) +
  coord_fixed()

# Display the plot
print(plot_matrix_c)
```

